---
title: "Identifying the Impacts of Extreme Weather"
format: docx
editor: visual
editor_options: 
  chunk_output_type: console
author: Joaquin Sandoval
date: 11/10/2025
---

Import necessary packages

```{r, message = FALSE}
library(tidyverse)
library(dplyr)
library(sf)
library(terra)
library(stars)
library(tmap)
library(kableExtra)
library(patchwork)
```

## Data Import

```{r, message = FALSE, output = FALSE}
# Import h08v05 tile February 7, 2021 
h08v05_2021_02_07 <- rast(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

# Import h08v05 tile February 7, 2021 STARS 
h08v05_2021_02_07_stars <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))

# Import h08v06 tile February 7, 2021
h08v06_2021_02_07 <- rast(here::here("data", "VNP46A1",
"VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# Import h08v06 tile February 7, 2021 STARS 
h08v06_2021_02_07_stars <- read_stars(here::here("data", "VNP46A1",
"VNP46A1.A2021038.h08v06.001.2021039064329.tif"))

# Import h08v05 tile February 16, 2021
h08v05_2021_02_16 <- rast(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

# Import h08v05 tile February 16, 2021 STARS 
h08v05_2021_02_16_stars <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))

# Import h08v06 tile February 16, 2021
h08v06_2021_02_16 <- rast(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Import h08v06 tile February 16, 2021 STARS 
h08v06_2021_02_16_stars <- read_stars(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

# Import Houston road data
roads <- st_read(here::here("data", "gis_osm_roads_free_1.gpkg"), 
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'")

# Import Houston houses data 
houses <- st_read(here::here("data", "gis_osm_buildings_a_free_1.gpkg"), 
                  query = "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)")

# Import data from from the U.S. Census Bureau’s American Community Survey for census tracts in 2019

acs_geom <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS")

acs_income <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "X19_INCOME")

houston_lines <- st_read(here::here("data", "COH_City_Limit_view_1951759729060774326", "Houston_City_Limit.shp"))

```

## Create blackout mask

Identify areas that experienced a blackout

```{r}
# Creating raster object of each day 

# Merging h08v05 and h08v06 tiles for February 7, 2021

houston_feb7 <- merge(h08v05_2021_02_07, h08v06_2021_02_07)

# Merging h08v05 and h08v06 tiles for February 16, 2021
houston_feb16 <- merge(h08v05_2021_02_16, h08v06_2021_02_16)

# Creating a difference raster; finding the change in night light intensity

difference_raster = houston_feb7 - houston_feb16 
```

Reclassify the difference raster. Any location that experienced a drop of more than 200 nW cm-2sr-1 experienced a blackout.

```{r}
# Creating a reclassification matrix 

# Assigning NA to locations experiencing < 200 nW cm-2sr-1 change
# Assigning value of 1 to locations experiencing >= 200 nW cm-2sr-1 change

rcl <- matrix(c(-Inf, 200, NA, 
                200, Inf, 1),
              ncol = 3, byrow = TRUE)

# Apply matrix to raster to reclassify: all cells are NA or 1 now

reclassified <- classify(difference_raster, rcl = rcl)

# Change reclassified values into factors 

values(reclassified) <- as.factor(values(reclassified))

```

Crop (spatially subset) the blackout mask to the Houston area as defined by the following coordinates: (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)

```{r}
# Creating an extent for the Houston area with the above coordinates

extent2 <- rast(xmin = -96.5, # xmin 
              xmax = -94.5, # xmax
              ymin = 29,    # ymin 
              ymax = 30.5) # ymax


# Applying the above defined extent to the reclassified raster 

cropped <- crop(reclassified, extent2)

# Very 'cropped' has fewer values 

if(ncell(reclassified) == nrow(cropped)) {
  warning("clipping did not remove cells")
} else {
  print("clipping removed cells")
}

```

Re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)

```{r}
# Reproject to EPSG 3083

reprojected_houston <- project(cropped,"EPSG:3083")

# Check if crs was updated 

if(st_crs(reprojected_houston)$epsg == 3083) {
  print("CRS was updated")
} else {
  print("CRS was not updated")
}

```

Vectorize raster

```{r}
# Convert raster data to polygon vector data 

cropped_blackout <- as.polygons(reprojected_houston) |> 
  st_as_sf()

# Confirm `cropped_blackout` is valid sf object 

st_is_valid(cropped_blackout)

```

## Exclude highways from the cropped blackout mask

```{r}
# Transform CRS of roads to match blackout map with updated CRS 

highway_transform <- st_transform(roads, crs = st_crs(cropped_blackout))
# Using if-else statements to verify match
if(st_crs(highway_transform) == st_crs(cropped_blackout)){
print("Match")
} else {
print("Not a match")
}

```

```{r}
# Checking that units are the same 

if (st_crs(cropped_blackout, parameters = TRUE)$units == st_crs(highway_transform, parameters = TRUE)$units) {
  print("Units match") 
} else {
  print("Units do not match")
}

```

Exlude areas within 200m of highways.

```{r}
# Creating a 200m buffer from highways 

highway_200m_buffer <- st_buffer(highway_transform, dist = 200) |>
  st_union()
```

```{r, warning=FALSE}
# Finding the difference between the blackout map and the highway buffer 
# Returns cropped_black out that does not overlay with the buffer 

difference <- st_difference(cropped_blackout, highway_200m_buffer)

```

```{r}

tm_shape(difference) + 
  tm_polygons(fill = "black") + 
  tm_title("Houston Areas Affected by Blackout", 
          fontface = "bold") + 
  tm_scalebar(position = c("left", "bottom")) + 
  tm_compass(position = c("left", "bottom")) + 
  tm_basemap("OpenStreetMap")

```

## Set of maps comparing night light intensities before and after the first two storms

```{r}
# Creating stars raster object of each day 

# Merge tils for February 7  
houston_feb7_stars <- st_mosaic(h08v05_2021_02_07_stars, h08v06_2021_02_07_stars )

# Merge tiles for February 16 
houston_feb16_stars <- st_mosaic(h08v05_2021_02_16_stars, h08v06_2021_02_16_stars)
# Define the new extent

extent <- st_bbox(c(xmin = -96.5, # xmin
                  ymin = 29, # ymin 
                  xmax = -94.5, # xmax
                  ymax = 30.5), 
                  crs = st_crs(houston_feb7_stars))

# Crop the Feb 7 stars object
cropped_houston_feb7_stars <- st_crop(houston_feb7_stars, extent) 

# Crop Feb 16 stars object
cropped_houston_feb16_stars <- st_crop(houston_feb16_stars, extent) 

plot(cropped_houston_feb7_stars, 
     main = "February 7, 2021",
     key.pos = NULL)
plot(cropped_houston_feb16_stars, 
     main = "February 16, 2021",
     key.pos = NULL)

```

## Identify the number of homes likely impacted by blackouts

```{r}
# Transform CRS of houses to match the difference crs

houses_transform <- st_transform(houses, crs = st_crs(difference))
# Using if-else statements to verify match
if(st_crs(houses_transform) == st_crs(difference)){
print("Match")
} else {
print("Not a match")
}
```

```{r}
# Checking that units are the same 

if (st_crs(houses_transform, parameters = TRUE)$units == st_crs(difference, parameters = TRUE)$units) {
  print("Units match") 
} else {
  print("Units do not match")
}

```

```{r}
# Make both objects valid prior to doing st_intersection

houses_transform <- houses_transform |> 
  st_make_valid()

difference <- difference |> 
  st_make_valid()

houses_blackout <- st_intersects(houses_transform, difference)

# Convert to logical vector

houses_blackout_logical <- lengths(houses_blackout) > 0

# filter based on logical vector

house_blackout = houses_transform[houses_blackout_logical, ]
```

```{r}
# Finding number of homes affected by blackout 

house_blackout |> 
  summarise(n_homes = n()) |> 
  st_drop_geometry()

```

## Socioecomomic: distributions of median household income for census tracts that did and did not experience blackouts.

```{r}
# Left join between layers with GEO ID to join income layer with geometry layer

acs_sf_df <- acs_geom |> 
  left_join(acs_income, by = c("GEOID_Data" = "GEOID"))
```


```{r}
# Transform CRS of acs_sf_df to match the difference crs

acs_sf_df_transform <- st_transform(acs_sf_df, crs = st_crs(difference))
# Using if-else statements to verify match
if(st_crs(acs_sf_df_transform) == st_crs(difference)){
print("Match")
} else {
print("Not a match")
}
```

```{r, message=FALSE, output = FALSE}
# Make acs_sf_df_transform valid 
acs_sf_df_transform |> 
  st_make_valid()
```

```{r}
tracts_within_blackout <- st_intersects(acs_sf_df_transform, difference)

# Convert to logical vector
tracts_within_blackout_logical <- lengths(tracts_within_blackout) > 0

# filter based on logical vector, only have buildings in blackout
income_blackout = acs_sf_df_transform[tracts_within_blackout_logical, ]

```

```{r}
# Boxplot income within blackout 

plot1 <- ggplot(income_blackout) + 
  geom_boxplot(aes(y = B19013e1)) + 
  labs(title = "Inside Blackout Zone", 
       y = "Household Income") + 
  theme(axis.ticks.x = element_blank(), 
        axis.text.x = element_blank()) 

```

```{r, warning = FALSE}
# Finding census tracts outside of blackout 

tracts_out_blackout <- st_difference(difference, acs_sf_df_transform)
```

```{r, warning=FALSE}
# Boxplot income outside blackout 
plot2 <- ggplot(tracts_out_blackout)  +
  geom_boxplot(aes(y = B19013e1)) + 
  labs(title = "Outside Blackout Zone") + 
  theme(axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank()) 
```

```{r, warning=FALSE}
plot1 + plot2
```

##  Map of the census tracts in Houston that lost power

```{r}
# Plot the tracts that overlapped with blackout mask `income_blackout`

tm_shape(income_blackout) + 
  tm_polygons(fill = "lightgreen") +
tm_title("Houston Census Tracts Affected by Blackout", 
          fontface = "bold") + 
  tm_scalebar(position = c("left", "bottom")) + 
  tm_compass(position = c("left", "bottom")) + 
  tm_basemap("OpenStreetMap")
```

## Interpretation 

The above analysis revealed the extent to which the Houston area experienced power outages after three severe winter storms in February of 2021. This analysis 
was based on data from Visible Infrared Imaging Radiometer Suite (VIIRS) onboard the Suomi satellite. Areas that experienced a loss of more than than 200 nW cm-2sr-1 (radiant intensity) were assigned to a blackout mask. This mask excluded areas 200 meters from highways. An estimated 133,788 buildings were in this area that lost power. Further analysis was conducted to determine differences in median household income between areas within this blackout area and those not. As the boxplot shows, the median income between both groups is approximately showing that this is power outage caused by extreme weather events did not discriminate based on wealth. 
